---
title: "Uncertainty in the prevalance of ecological chaos"
author: "Erik Tihelka"
date: '2022-07-09'
output: html_document
---

Re-examining the prevalence of chaos in ecological time series, after Rogers et al. (*Nat. Ecol. Evol.*, 2022), while accounting for uncertainty in false positive and false negative rates of chaos detection methods.

## Initial Setup

```{r, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
library(rstan)
library(rstudioapi)
options(mc.cores = parallel::detectCores())
```

## Shortest posterior interval

Define the shortest posterior interval, which is more sensible than a central interval because of the skewness of the posterior and the hard boundary at 0. It is equivalent to the highest posterior density interval for unimodal posteriors.

Code and commentary adapted from Gelman & Carpenter (*Appl. Statist.*,
2020, pp. 1269–1283).

```{r}
spin <- function(x, lower=NULL, upper=NULL, conf=0.95){
  x <- sort(as.vector(x))
  if (!is.null(lower)) {
    if (lower > min(x)) stop("lower bound is not lower than all the data")
    else x <- c(lower, x)
  }
  if (!is.null(upper)) {
    if (upper < max(x)) stop("upper bound is not higher than all the data")
    else x <- c(x, upper)
  }
  n <- length(x)
  gap <- round(conf*n)
  width <- x[(gap+1):n] - x[1:(n-gap)]
  index <- min(which(width==min(width)))
  x[c(index, index + gap)]
}
```

## Bayesian analysis

Build a simple Bayesian model where the false positive $f_p$ and false negative $f_n$ rates are uncertain. We will use a beta distribution to model uncertainty of the both false positive and false negative probabilities. 

```{r}
stan.input2 <- function(y=58, n=172, alpha=1, beta=1, an=9.35, bn=22.40,  ap=0.28, bp=1.51, nchains=4)
{
  inits <- list()
  bugs.data <- list(y=y, n=n, alpha=alpha, beta=beta, an=an, bn=bn, ap=ap, bp=bp)
  for (i in 1:nchains) 
    inits[[i]] <- list(theta=runif(1), fn=runif(1), fp=runif(1))
  para <- c("theta", "fn", "fp")
  return(list(para=para, data=bugs.data, inits=inits, n.chains=nchains, model = "
         data{
          int n;
          int y;
          real alpha;
          real beta;
          real an;
          real bn; 
          real ap;
          real bp; 
         }
         parameters{
          real<lower=0,upper=1> theta;
          real<lower=0,upper=1> fn;  
          real<lower=0,upper=1> fp;  
         }
         model {
          fn ~ beta(an, bn);
          fp ~ beta(ap, bp);
          theta ~ beta(alpha, beta);
          y ~ binomial(n, theta*(1-fn)+(1-theta)*fp);
         }
         "))
}
input.to.stan <- stan.input2()
fitB <- stan(model_code = input.to.stan$model, data=input.to.stan$data, 
            init = input.to.stan$inits, pars = input.to.stan$para, 
            iter = 1, chains = input.to.stan$n.chains)
fitB2 <- stan(fit=fitB, data = input.to.stan$data,
             init=input.to.stan$inits, pars=input.to.stan$para,
             iter = 500000, chains = input.to.stan$n.chains, thin=700)
print(fitB2)
```

## 95% intervals

Calculates the shortest posterior interval of $\theta$.

```{r}
draws_1 = extract(fitB2, pars=c("theta"))
print(spin(draws_1[["theta"]], lower=0, upper=1, conf=0.95))
```

We evaluated the uncertainty on the prevalence of chaos as detected with three different chaos detection methods in Rogers et al. (*Nat. Ecol. Evol.*, 2022). Our results are reported in Supplementary Table 1. Likewise, for the justification of prior choice, see our Supplementary Discussion.
